---
title: "Introduction to multiCollglm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to multiCollglm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`multiCollglm` diagnoses collinearity in generalized linear models (GLMs),
generalizing the classical linear-regression diagnostics (condition
number, Belsley-Kuh-Welsch decomposition, and the Redefined Variance
Inflation Factor, RVIF [@salmeron2025]) to any family and link function. All three are
computed on the design matrix weighted by the IRLS weights at
convergence and, by default, scaled to unit length by column (without
centering). `ml_collinearity()` adds a fourth, complementary diagnostic:
Lesaffre and Marx's [-@lesaffre1993] comparison between the condition
number of the *original* design matrix and that of the IRLS-weighted
information matrix, used to tell apart ordinary collinearity among the
explanatory variables from *ML-collinearity* (see below). This vignette
is the package's complete guide; see `vignette("multiCollglm")` any time
you need to come back to it.

## Installation

```r
# install.packages("remotes")
remotes::install_github("cbgarciaugr/multiCollglm")
```

## Usage with `glm`

```{r}
library(multiCollglm)

set.seed(1)
n <- 200
x1 <- rnorm(n)
x2 <- x1 + rnorm(n, sd = 0.05) # almost collinear with x1
x3 <- rnorm(n)
x4 <- rnorm(n)
mu <- exp(1 + 0.3 * x1 + 0.3 * x2 - 0.2 * x3 + 0.1 * x4)
y <- rgamma(n, shape = 5, rate = 5 / mu)

mod <- glm(y ~ x1 + x2 + x3 + x4, family = Gamma(link = "inverse"))
```

```{r}
condition_number(mod)
```

```{r}
bkw_diagnostics(mod)
```

```{r}
rvif_diagnostics(mod)
```

## Condition-number methods

`condition_number()` also accepts a `method` argument that switches
between five published (or, for `"RAW"`, textbook-baseline) definitions
of CN for a GLM:

```{r}
condition_number(mod, method = "RAW") # no transformation at all
condition_number(mod, method = "MP")  # requires multiColl package
condition_number(mod, method = "WS")  # same as the default above, explicitly labeled
condition_number(mod, method = "OZ")  # centered and unit-scaled [@ozkale2021]
condition_number(mod, method = "MS")  # centered, unit-scaled, then refit [@marx1990]
```

| `method` | Transformation | Comments |
|---|---|---|
| `"RAW"` | None at all: eigenvalues of $X'\hat{W}X$ directly (intercept included, no centering, no scaling). The plain baseline against which the other four can be judged -- usually dominated by however differently-scaled the explanatory variables happen to be. |  |
| `"MP"` | @mackinnon1989. The *original* explanatory variables (intercept included) are rescaled to unit Euclidean length with [`multiColl::lu()`](https://cran.r-project.org/package=multiColl) **before** fitting. | A GLM's IRLS weights at convergence are invariant to any linear rescaling of `X` -- but done explicitly to mirror the original definition. |
| `"MS"` | @marx1990. The explanatory variables (excluding the intercept, if any) are **centered on their own mean and rescaled to unit Euclidean length before fitting**, with an ordinary intercept re-added if the original model had one. | Unlike `"MP"`, centering shifts the fitted values themselves, so the refit is genuinely different from `model`. |
| `"WS"` | @weissfeld1991, the package default. The GLM is fit on the original variables and the IRLS-weighted information matrix is scaled to unit column length *afterwards* (no centering). |  |
| `"OZ"` | @ozkale2021. Same as `"WS"`, but the IRLS-weighted design matrix is additionally **centered** before being scaled to unit length. If `model` has an intercept, its column is **dropped** before centering/scaling (every worked example in Ozkale 2019, Sec. 5, fits without an intercept term). |  |

A few things worth knowing before picking one:

- **`"MP"` requires the `multiColl` package** (`Suggests`, not a hard
  dependency); it errors with an informative message if it isn't
  installed.
- **`"MS"` can fail to converge for a model with no intercept.**
  Centering removes every column's mean. For some family/link
  combinations (most notably Gamma with the inverse link, which needs
  strictly positive fitted values) the refit's IRLS iteration can
  diverge. When that happens you get an informative error explaining why
  this is expected for that combination and is not a bug: @marx1990
  developed the method for logistic regression models, which ordinarily
  do include an intercept. If you hit this, use `"WS"` or `"OZ"` for that
  particular model instead.
- **Centering an IRLS-weighted design matrix that includes an intercept
  can leave it exactly rank-deficient** for *some* family/link
  combinations (for example the Gamma family with its canonical
  (inverse) link specifically -- `sqrt(w) * eta` is exactly 1 for every
  observation, regardless of whether an intercept is present). `"OZ"`
  sidesteps this for most families by dropping the intercept before
  centering, but for Gamma-inverse the singularity is structural either
  way and `NC_OZ` will then legitimately come back as `Inf` with a
  warning. See the [`Nitrogen` reproduction
  article](https://cbgarciaugr.github.io/multiCollglm/articles/reproduction-nitrogen.html)
  for a worked example.
- Leave `method = NULL` (the default) to keep using `center`/`scale`
  directly, exactly as in earlier versions of this function -- see the
  next section.

`"MP"` [@mackinnon1989] additionally requires the
`multiColl` package. See `?condition_number` for the full argument
reference.

## Changing the data transformation

All three functions accept `center` and `scale` arguments, applied to the
IRLS-weighted design matrix before computing any diagnostic (ignored
whenever `method` is supplied). They accept the same values as base R's
own `scale()`:

- `center`: `FALSE` (default), `TRUE` (subtract each column's mean), or a
  numeric vector of per-column values to subtract.
- `scale`: `"unit"` (default; Euclidean/L2 unit-length columns, the
  originally proposed transformation), `TRUE` (root-mean-square scaling),
  `FALSE` (no scaling), or a numeric vector of per-column divisors.

```{r}
condition_number(mod, scale = TRUE) # root-mean-square scaling instead of unit length
```

`scale = TRUE` rescales every eigenvalue by the same factor (`nrow(X) -
1`) relative to the default `scale = "unit"`, so `CN` and `sqrt(CN)` are
unaffected; `scale = FALSE` or a custom numeric vector generally do
change them, since the columns are no longer put on a comparable scale.

**Do not use `center = TRUE`** on an intercept model unless you know what
you are doing: for a canonical-link GLM, the IRLS-weighted design matrix
satisfies `Xw %*% coef(mod) == 1` exactly, so centering the intercept
column is guaranteed to drop the matrix's rank by exactly one. This is
precisely why Belsley, Kuh and Welsch [-@belsley1980] recommend against centering
for collinearity diagnostics; `bkw_diagnostics()` and `rvif_diagnostics()`
detect this and raise an informative error instead of returning `NaN`.

```{r, error = TRUE}
bkw_diagnostics(mod, center = TRUE)
```

## Usage with `glmnet` / `cv.glmnet`

When the model is a regularized fit (`glmnet` or `cv.glmnet`), there is no
IRLS fit and no convergence weights of its own. `multiCollglm` identifies
the active set of variables at the given `lambda` and refits a standard
`glm()` on them so the same diagnostics can be computed:

```r
library(glmnet)

x <- as.matrix(data.frame(x1, x2, x3, x4))
cvfit <- cv.glmnet(x, y, family = Gamma(link = "inverse"))

condition_number(cvfit, x = x, y = y, family = Gamma(link = "inverse"), s = "lambda.min")
bkw_diagnostics(cvfit, x = x, y = y, family = Gamma(link = "inverse"), s = "lambda.min")
rvif_diagnostics(cvfit, x = x, y = y, family = Gamma(link = "inverse"), s = "lambda.min")
```

The result also includes `active_vars` and `dropped_vars`, so you know
exactly which sub-model the diagnostic was computed on.

## BKW rule of thumb

`bkw_diagnostics()` flags as suspicious any dimension whose `sqrt(CN)`
is `>= 10` and on which at least two coefficients have a variance
proportion `>= 0.5` [@belsley1980]. Both thresholds are
adjustable via `index_threshold` and `proportion_threshold`.

## ML-collinearity

`ml_collinearity()` implements the diagnostic of Lesaffre and Marx
[-@lesaffre1993], which compares `sqrt(CN)` of the *original* design
matrix (via `multiColl::CNs()`) against `sqrt(CN)` of the IRLS-weighted
information matrix under `method = "MP"`, to distinguish ordinary
collinearity among the explanatory variables from *ML-collinearity*:
ill-conditioning that instead arises from the combination of the
response, the fitted coefficients and the link function.

```{r}
ml_collinearity(mod)
```

`kappa_x` (`sqrt(CN)` of `X`) is obtained via
[`multiColl::CNs()`](https://cran.r-project.org/package=multiColl):
`CNs(X)` returns a list with `CN1` ("Condition Number without
intercept") and `CN2` ("Condition Number with intercept"); since
`kappa_X` standardizes `X` *including* the constant vector, `CNs(X)$CN2`
is the one used.

```
ML-collinearity diagnostic (Lesaffre and Marx, 1993)

sqrt(CN) of X (original design matrix, multiColl::CNs(X)$CN2): 190.7765
sqrt(CN) of W (MacKinnon-Puterman information matrix, method = "MP"): 329.9542
ratio (sqrt(CN)_W / sqrt(CN)_X): 1.7295

Collinearity in X (sqrt(CN)_X > 30): YES
ML-collinearity (ratio_wx > 5 and sqrt(CN)_W > 30): no
```

(the printed example above is Lesaffre and Marx's own Lee
cancer-remission model, reproduced exactly in the [`LeeCancer`
reproduction
article](https://cbgarciaugr.github.io/multiCollglm/articles/reproduction-lee-cancer.html).)

Following Lesaffre and Marx (1993, Sec. 4.2), a model is flagged with
`ml_collinearity = TRUE` only when **both** the ratio `sqrt(CN)_W /
sqrt(CN)_X` exceeds `ratio_threshold` (default 5) **and** `sqrt(CN)_W`
itself exceeds `kappa_threshold` (default 30) -- a high ratio by itself,
with a small `sqrt(CN)_W`, does not indicate an ill-conditioned
information matrix to begin with. Separately, `sqrt(CN)_X >
kappa_threshold` flags ordinary collinearity among the explanatory
variables (`collinearity_x`); if both flags are set and the ratio is
close to 1, both types of collinearity are simultaneously present.
`ml_collinearity()` also has `glmnet`/`cv.glmnet` methods, refitting on
the active set exactly as the other three diagnostics do. **Requires
the `multiColl` package** (`Suggests`, not a hard dependency). See
`?ml_collinearity` for the full argument reference.

## Beyond this guide

This vignette covers the package's basic usage. The [package
website](https://cbgarciaugr.github.io/multiCollglm/articles/index.html)
also includes a series of **reproduction articles**: real cases from the
literature where collinearity was diagnosed without applying the correct
data transformation, together with a comparison against the results
`multiCollglm` produces.

## References
