---
title: "Introduction to multiCollglm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to multiCollglm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`multiCollglm` diagnoses collinearity in generalized linear models (GLMs),
generalizing the classical linear-regression diagnostics (condition
number, Belsley-Kuh-Welsch decomposition, and the Redefined Variance
Inflation Factor, RVIF) to any family and link function. All three are
computed on the design matrix weighted by the IRLS weights at
convergence and, by default, scaled to unit length by column (without
centering).

## Installation

```r
# install.packages("remotes")
remotes::install_github("cbgarciaugr/multiCollglm")
```

## Usage with `glm`

```{r}
library(multiCollglm)

set.seed(1)
n <- 200
x1 <- rnorm(n)
x2 <- x1 + rnorm(n, sd = 0.05) # almost collinear with x1
x3 <- rnorm(n)
x4 <- rnorm(n)
mu <- exp(1 + 0.3 * x1 + 0.3 * x2 - 0.2 * x3 + 0.1 * x4)
y <- rgamma(n, shape = 5, rate = 5 / mu)

mod <- glm(y ~ x1 + x2 + x3 + x4, family = Gamma(link = "inverse"))
```

```{r}
condition_number(mod)
```

```{r}
bkw_diagnostics(mod)
```

```{r}
rvif_diagnostics(mod)
```

## Condition-number methods

`condition_number()` also accepts a `method` argument that switches
between five published (or, for `"RAW"`, textbook-baseline) definitions
of the condition number for a GLM:

```{r}
condition_number(mod, method = "RAW") # no transformation at all
condition_number(mod, method = "WS")  # same as the default above, explicitly labeled
condition_number(mod, method = "OZ")  # centered and unit-scaled (Ozkale, 2019)
condition_number(mod, method = "MS")  # centered, unit-scaled, then refit (Marx and Smith, 1990)
```

`"MP"` (Mackinnon and Puterman, 1989) additionally requires the
`multiColl` package. See `?condition_number` and the package's
[README](https://github.com/cbgarciaugr/multiCollglm#condition-number-methods-method-argument)
for the full description of each method, and for the caveats around
`"MS"` (which can fail to converge for a no-intercept model) and `"OZ"`
(which can legitimately return `Inf` for a Gamma model with the
canonical inverse link).

## Changing the data transformation

All three functions accept `center` and `scale` arguments, applied to the
IRLS-weighted design matrix before computing any diagnostic (ignored
whenever `method` is supplied). They accept the same values as base R's
own `scale()`:

- `center`: `FALSE` (default), `TRUE` (subtract each column's mean), or a
  numeric vector of per-column values to subtract.
- `scale`: `"unit"` (default; Euclidean/L2 unit-length columns, the
  originally proposed transformation), `TRUE` (root-mean-square scaling),
  `FALSE` (no scaling), or a numeric vector of per-column divisors.

```{r}
condition_number(mod, scale = TRUE) # root-mean-square scaling instead of unit length
```

**Do not use `center = TRUE`** on an intercept model unless you know what
you are doing: for a canonical-link GLM, the IRLS-weighted design matrix
satisfies `Xw %*% coef(mod) == 1` exactly, so centering the intercept
column is guaranteed to drop the matrix's rank by exactly one. This is
precisely why Belsley, Kuh and Welsch (1980) recommend against centering
for collinearity diagnostics; `bkw_diagnostics()` and `rvif_diagnostics()`
detect this and raise an informative error instead of returning `NaN`.

```{r, error = TRUE}
bkw_diagnostics(mod, center = TRUE)
```

## Usage with `glmnet` / `cv.glmnet`

When the model is a regularized fit (`glmnet` or `cv.glmnet`), there is no
IRLS fit and no convergence weights of its own. `multiCollglm` identifies
the active set of variables at the given `lambda` and refits a standard
`glm()` on them so the same diagnostics can be computed:

```r
library(glmnet)

x <- as.matrix(data.frame(x1, x2, x3, x4))
cvfit <- cv.glmnet(x, y, family = Gamma(link = "inverse"))

condition_number(cvfit, x = x, y = y, family = Gamma(link = "inverse"), s = "lambda.min")
bkw_diagnostics(cvfit, x = x, y = y, family = Gamma(link = "inverse"), s = "lambda.min")
rvif_diagnostics(cvfit, x = x, y = y, family = Gamma(link = "inverse"), s = "lambda.min")
```

The result also includes `active_vars` and `dropped_vars`, so you know
exactly which sub-model the diagnostic was computed on.

## BKW rule of thumb

`bkw_diagnostics()` flags as suspicious any dimension whose condition
index is `>= 10` and on which at least two coefficients have a variance
proportion `>= 0.5` (Belsley, Kuh and Welsch, 1980). Both thresholds are
adjustable via `index_threshold` and `proportion_threshold`.

## Beyond this guide

This vignette covers the package's basic usage. The [package
website](https://cbgarciaugr.github.io/multiCollglm/articles/index.html)
also includes a series of **reproduction articles**: real cases from the
literature where collinearity was diagnosed without applying the correct
data transformation, together with a comparison against the results
`multiCollglm` produces.

## References

- Belsley, D.A., Kuh, E. and Welsch, R.E. (1980). *Regression
  Diagnostics: Identifying Influential Data and Sources of
  Collinearity*. Wiley.
- Salmeron, R., Garcia, C.B. and Garcia, J. (2025). A redefined Variance
  Inflation Factor: overcoming the limitations of the Variance Inflation
  Factor. *Computational Economics*, 65, 337-363.
  <https://doi.org/10.1007/s10614-024-10575-8>
