---
title: "Reproduction: coal mine roof-fracture data (Mine)"
description: >
  Reproduces the collinearity diagnostic of the coal mine roof-fracture
  dataset, used by five works fitting different count-data GLMs, and
  shows that the "RAW" (untransformed) method exactly reproduces the
  most-cited value in the literature.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

The `Mine` dataset records 44 observations of roof fractures in coal
mines in the Appalachian region of West Virginia, with four continuous
predictors: inner burden thickness, percentage extraction of the
previously mined lower seam, lower seam height, and time since the mine
opened. It is the real-data application shared by five reviewed works,
each fitting a different count-data GLM: Kurtoglu and Ozkale (2019)
[`review55`] and Kurtoglu and Ozkale (2019) [`review118`] (same authors,
a restricted ridge and a restricted Liu estimator respectively, both with
Poisson and a log link); Sami, Amin and Butt (2022) [`review8`]
(Conway-Maxwell-Poisson, a "COMPRR" ridge estimator); Hussein and Algamal
(2026) [`review115`] (Bell regression, a Kibria-Lukman-type estimator);
and Sami et al. (2026) [`review117`] (discrete Weibull, a "DWRE" ridge
estimator).

`review8` states that it analyzes a different distribution but provides
no diagnostic measures of its own (which should differ across
distributions), instead citing a previous paper to claim that the dataset
exhibits collinearity. `review117` does the same, citing `review8` despite
also using a different distribution; moreover, based on the code in
`review117`'s appendix, the diagnostic appears to be computed on the
unweighted $X'X$, and although the code includes a standardization step,
it is never actually applied.

Note on the data: two transcription errors present in some circulating
copies of this dataset (row 30: `y = 22` instead of `y = 2`; row 36:
`x1 = 0` instead of `x1 = 80`) have been corrected in `data(Mine)` after
cross-checking against the appendix data table of a second paper.

## Data

```{r}
library(multiCollglm)
data(Mine)
str(Mine)

mod <- glm(y ~ x1 + x2 + x3 + x4, family = poisson(link = "log"),
           data = Mine)
summary(mod)
```

## Original diagnostic (a tour of the literature)

Kurtoglu and Ozkale (2019), fitting Poisson with a log link, obtain a
condition number of **2093.32** that at first could not be reproduced
with any of the package's centered or scaled transformations. Hussein and
Algamal (2026), with a Bell distribution, obtain **2139.89** (same order
of magnitude, consistent with a different but Poisson-like count
distribution for these data).

## Diagnostic with multiCollglm

```{r}
methods <- c("RAW", "MP", "MS", "WS", "OZ")

tbl <- do.call(rbind, lapply(methods, function(m) {
  res <- tryCatch(condition_number(mod, method = m),
                   error = function(e) NULL)
  if (is.null(res)) {
    data.frame(method = m, nc_label = NA, condition_number = NA,
               condition_index = NA)
  } else {
    data.frame(method = m, nc_label = res$nc_label,
               condition_number = res$condition_number,
               condition_index = res$condition_index)
  }
}))
knitr::kable(tbl, digits = 4)
```

## Comparison

The condition index (square-root scale) of `method = "RAW"` -- no
transformation at all, directly on $X'\hat{W}X$ -- is
**`r round(condition_number(mod, method = "RAW")$condition_index, 4)`**,
which matches, up to rounding, the **2093.32** reported by Kurtoglu and
Ozkale (2019) [`review55`, `review118`]. In other words: the value the
text initially describes as "not reproducible" **is exactly reproduced**
once the `"RAW"` method is added to the package -- what those two works
compute is the condition index (square root of the eigenvalue ratio) of
the IRLS-weighted information matrix, with no centering or scaling at
all.

When the data are rescaled to unit length without centering (`"WS"`,
which does include the intercept), the condition index drops to
**`r round(condition_number(mod, method = "WS")$condition_index, 2)`**
-- of the same order as the reference value (~28.8) obtained with this
transformation; the conclusion of severe collinearity holds, but with a
much more moderate and stable value, no longer dominated by the very
different scales of the four variables (thickness, percentage, height,
years).

We could not independently verify Hussein and Algamal's (2026) value for
the Bell regression, since that family is not part of R base's `glm()`
families and would require an additional package; it is left as reported
by the original source.

| | Literature | multiCollglm |
|---|---|---|
| Transformation | None (Poisson, log link, with intercept) | `"RAW"`: untransformed |
| Value (condition index, square root) | 2093.32 (Kurtoglu and Ozkale 2019) | `r round(condition_number(mod, method = "RAW")$condition_index, 4)` |
| Rescaled to unit length, no centering | -- | `r round(condition_number(mod, method = "WS")$condition_index, 4)` |
| Conclusion about collinearity | Severe | Severe with `"RAW"`; still relevant but less extreme with `"WS"` |

## Conclusion

This dataset is a good example of why it's worth including `method =
"RAW"` as an explicit baseline: the most-cited reference value for the
mine data turns out to be exactly the condition index with no
transformation whatsoever -- not an error or an undocumented calculation,
but simply the plainest possible case, which wasn't among this package's
published methods until it was added. Compared with `"WS"` (rescaled to
unit length, no centering), the diagnostic's magnitude changes
substantially because the four regressors are on very different units and
`"RAW"` is dominated by that scale disparity, not only by the underlying
collinearity.

## References

- Kurtoglu, F. and Ozkale, M.R. (2019). Restricted ridge estimator in
  generalized linear models: Monte Carlo simulation studies on Poisson
  and binomial distributed responses. *Communications in Statistics -
  Simulation and Computation*, 48(4), 1191-1218.
  <https://doi.org/10.1080/03610918.2017.1408822>
- Kurtoglu, F. and Ozkale, M.R. (2019). Restricted Liu estimator in
  generalized linear models: Monte Carlo simulation studies on gamma and
  Poisson distributed responses. *Hacettepe Journal of Mathematics and
  Statistics*, 48(4), 1250-1276. <https://doi.org/10.15672/HJMS.2018.624>
- Sami, F., Amin, M. and Butt, M.M. (2022). On the ridge estimation of the
  Conway-Maxwell Poisson regression model with multicollinearity: Methods
  and applications. *Concurrency and Computation: Practice and
  Experience*, 34(1), e6477. <https://doi.org/10.1002/cpe.6477>
- Hussein, S.M.J.A. and Algamal, Z.Y. (2026). A New Shrinkage Estimator for
  the Bell Regression Model. *Thailand Statistician*, 24(2), 462-473.
- Sami, F., Aljohani, H.M., Iqbal, K., Ghorashi, S. and Abu Bakar Siddique,
  M. (2026). A new estimator for discrete Weibull regression model with
  correlated variables: an illustrative example. *Journal of Statistical
  Computation and Simulation*, 96(11), 2737-2757.
  <https://doi.org/10.1080/00949655.2026.2636232>
- Myers, R.H. (1990). *Classical and Modern Regression with
  Applications* (2nd ed.). Duxbury Press.
