---
title: "Reproduction: cancer remission data (Lee 1974)"
description: >
  Reproduces the three logistic-model variants fitted to Lee's (1974)
  cancer remission dataset -- Lesaffre and Marx (1993), Ozkale and Arican
  (2016)/Ozkale (2021), and Huang, Jou and Cho (2015) -- and contrasts
  them against multiCollglm's five methods.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

Lee's (1974) cancer remission dataset records clinical observations from
27 patients, of whom 9 had complete remission (`y = 1`) and 18 had
incomplete remission (`y = 0`), together with six quantitative
covariates: cellularity index (CELL), smear index (SMEAR), infiltrate
index (INFIL), labeling index (LI), percentage of blast cells (BLAST) and
body temperature (TEMP). It is a classic example for collinearity
diagnostics in binomial logistic regression, used with different subsets
of covariates by Lesaffre and Marx (1993) [`review133`], Ozkale and
Arican (2016) [`review35`] together with Ozkale (2021) [`review42`], and
Huang, Jou and Cho (2015) [`review137`].

Lesaffre and Marx (1993) are the first to use this dataset, with only
three covariates (LI, TEMP, CELL) that show severe collinearity -- a
classic scenario of extreme variance inflation and numerical instability.
The remaining reviewed works use all covariates except, in some cases,
LI.

## Data

```{r}
library(multiCollglm)
data(LeeCancer)
str(LeeCancer)
# TEMP is recorded in the original source on a /100 scale; it is
# multiplied back by 100 to match the scale used by the cited works.
LeeCancer$x6100 <- LeeCancer$x6 * 100
```

## Model 1: Lesaffre and Marx (1993) -- CELL, LI, TEMP

```{r}
mod133 <- glm(y ~ x1 + x4 + x6100, family = binomial(), data = LeeCancer)
summary(mod133)
```

## Model 2: Ozkale and Arican (2016) / Ozkale (2021) -- CELL, SMEAR, INFIL, BLAST, TEMP (no LI)

```{r}
mod3542 <- glm(y ~ x1 + x2 + x3 + x5 + x6100, family = binomial(),
                data = LeeCancer)
summary(mod3542)
```

Ozkale and Arican (2016) standardize and center the explanatory variables
before fitting, explicitly stating that *"standardization removes any
non-essential ill-conditioning resulting from the intercept"* -- that is,
they compute a **classical condition number, without any IRLS
weighting**, on the standardized data. They obtain $CN = 295.7026$
(eigenvalue ratio, no square root), which on the square-root scale would
be $17.196$.

Ozkale (2021), with the same model, reports the untransformed eigenvalues
$15.6562, 0.4601, 0.1557, 0.0145, 0.0001$ and from those a condition
number of $1.3032\times10^{5}$ -- a value that is obtained exactly neither
as the ratio ($15.6562/0.0001 = 156,562$) nor as its square root
($\approx 395.7$) from those very eigenvalues she reports. The same
happens with her centered-and-scaled scenario: eigenvalues $4.4133,
0.3657, 0.2009, 0.0200, 0.0002$, from which the ratio would be
$22,066.5$, not the $2.6099\times10^{4}$ she reports.

This same model (5 covariates, no LI, no intercept) was the subject of an
additional check earlier in this project: refitting the model from four
different starting points (least squares, zeros, sign-reversed least
squares, and `glm()`'s own defaults), all four converge to the same
unique point $(38.9501, 46.4911, -50.9001, 2.2396, -38.3061)$ -- which,
since logistic regression is a convex problem, confirms it is the global
maximum-likelihood estimate for these data. That point does **not**
match the coefficients published by Ozkale (2019, Sec. 5.2) for the same
model ($-45.5605, 46.2072, -47.9364, 3.5485, -48.5129$), which suggests an
error in Ozkale's own paper rather than in the data (independently
verified against SAS documentation and against the original SAS
`datalines` for the `Remission` dataset).

## Model 3: Huang, Jou and Cho (2015) -- all covariates

```{r}
mod137 <- glm(y ~ x1 + x2 + x3 + x4 + x5 + x6100, family = binomial(),
              data = LeeCancer)
summary(mod137)
```

Huang, Jou and Cho (2015) use all six covariates and report $CN =
418.13$.

## Diagnostic with multiCollglm

```{r}
methods <- c("RAW", "MP", "MS", "WS", "OZ")

calc_table <- function(mod) {
  do.call(rbind, lapply(methods, function(m) {
    res <- tryCatch(condition_number(mod, method = m),
                     error = function(e) NULL)
    if (is.null(res)) {
      data.frame(method = m, nc_label = NA, condition_number = NA,
                 condition_index = NA)
    } else {
      data.frame(method = m, nc_label = res$nc_label,
                 condition_number = res$condition_number,
                 condition_index = res$condition_index)
    }
  }))
}

cat("Model 1 (Lesaffre and Marx), with intercept:\n")
knitr::kable(calc_table(mod133), digits = 4)

cat("Model 2 (Ozkale), with intercept:\n")
knitr::kable(calc_table(mod3542), digits = 4)
mod3542b <- glm(y ~ x1 + x2 + x3 + x5 + x6100 - 1, family = binomial(),
                 data = LeeCancer)
cat("Model 2 (Ozkale), without intercept (their own specification):\n")
knitr::kable(calc_table(mod3542b), digits = 4)

cat("Model 3 (Huang, Jou and Cho), with intercept:\n")
knitr::kable(calc_table(mod137), digits = 4)
mod137b <- glm(y ~ x1 + x2 + x3 + x4 + x5 + x6100 - 1, family = binomial(),
               data = LeeCancer)
cat("Model 3 (Huang, Jou and Cho), without intercept:\n")
knitr::kable(calc_table(mod137b), digits = 4)
```

## Comparison

For Model 2 (Ozkale), the classical, unweighted condition number reported
by Ozkale and Arican (2016), $CN = 295.7026$, does not match any of
`condition_number()`'s five methods, because these always weight by the
IRLS weights at convergence -- exactly the transformation that, per this
package's whole premise, should be applied and that most reviewed works
omit.

For Model 3 (Huang, Jou and Cho), the reported value ($418.13$) is
closest, in order of magnitude, to the condition index (square-root
scale) of `"MP"`/`"WS"` **without** an intercept
(`r round(condition_number(mod137b, method = "WS")$condition_index, 2)`
and `r round(condition_number(mod137b, method = "MP")$condition_index, 2)`
respectively) than to any of the with-intercept variants, but it does not
match exactly.

| | Literature | multiCollglm |
|---|---|---|
| Model 2, stated transformation | Centered and scaled, **not** IRLS-weighted | `"OZ"` (centered+scaled, IRLS-weighted, no intercept): `r round(condition_number(mod3542b, method = "OZ")$condition_index, 2)` |
| Model 3, transformation | Not specified | `"WS"` without intercept: `r round(condition_number(mod137b, method = "WS")$condition_index, 2)` (closest to the published 418.13) |

## Conclusion

The three published models fitted to this dataset each illustrate, in
their own way, why the transformation matters: Lesaffre and Marx (1993)
work with only three already-known-to-be-collinear covariates; Ozkale and
Arican (2016) compute a classical diagnostic without IRLS weighting that,
by construction, cannot match any of this package's methods (all of which
weight by the IRLS weights); and the number Ozkale (2021) reports from
her own eigenvalues shows an internal arithmetic inconsistency we could
not resolve. In addition, for the 5-covariate no-intercept model, the fit
Ozkale (2019) reports does not match the unique global maximum-likelihood
estimate for these data (verified by convergence from four different
starting points), which points to an error in her paper rather than in
the data, already independently verified against SAS's official
documentation.

## References

- Lee, E.T. (1974). A computer program for linear logistic regression
  analysis. *Computer Programs in Biomedicine*, 4(2), 80-92.
  <https://doi.org/10.1016/0010-468X(74)90011-7>
- Lesaffre, E. and Marx, B.D. (1993). Collinearity in Generalized Linear
  Regression. *Communications in Statistics - Theory and Methods*, 22(7),
  1933-1952. <https://doi.org/10.1080/03610929308831126>
- Ozkale, M.R. and Arican, E. (2016). A new biased estimator in logistic
  regression model. *Statistics*, 50(2), 233-253.
  <https://doi.org/10.1080/02331888.2015.1123711>
- Ozkale, M.R. (2021). The red indicator and corrected VIFs in
  generalized linear models. *Communications in Statistics - Simulation
  and Computation*, 50(12), 4144-4170.
  <https://doi.org/10.1080/03610918.2019.1639740>
- Huang, C.-C.L., Jou, Y.-J. and Cho, H.-J. (2015). A New Multicollinearity
  Diagnostic for Generalized Linear Models. *Journal of Applied
  Statistics*. <https://doi.org/10.1080/02664763.2015.1126239>
