---
title: "Reproduction: [Author(s), Year] -- [short title of the work]"
description: >
  Reproduces the collinearity diagnostic from [short citation] and
  compares it against the result from multiCollglm, which applies the
  correct transformation (design matrix weighted by IRLS, scaled to unit
  length, without centering).
bibliography: ../references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

<!--
TEMPLATE for a reproduction article.

How to use it:
1. Duplicate this file inside vignettes/articles/ with a descriptive
   name, e.g. reproduction-perez-2019-energy-consumption.Rmd
2. Replace the title/description in the YAML header above.
3. Replace the "Data" section with the real data from the work (or a
   faithful simulation if the original data cannot be redistributed:
   check its license first).
4. Replace the "Original diagnostic" section with the calculation
   exactly as it appears in the published work (or as it would be
   calculated following its methodological description), making clear
   what transformation (or lack of one) was applied.
5. Add the article to _pkgdown.yml (articles section) so it appears in
   the package website's menu.
6. Add any new references you cite (with @key) to vignettes/references.bib.
7. Delete this comment block before publishing.
-->

## Context

[2-3 sentence summary of the original work: what GLM they fitted, which
variables, and what conclusion they drew about collinearity. Full
citation in the References section at the end.]

## Data

```{r}
# TODO: replace with the real data (or a faithful simulation) from the
# original work. Illustrative example with simulated data:
set.seed(1)
n <- 200
x1 <- rnorm(n)
x2 <- x1 + rnorm(n, sd = 0.05) # almost collinear with x1
x3 <- rnorm(n)
mu <- exp(1 + 0.3 * x1 + 0.3 * x2 - 0.2 * x3)
y <- rgamma(n, shape = 5, rate = 5 / mu)

mod <- glm(y ~ x1 + x2 + x3, family = Gamma(link = "inverse"))
summary(mod)
```

## Original diagnostic (without the correct transformation)

[Explain what transformation (or lack of one) the original work applied:
for example, classical VIF on the unweighted matrix, or data centered
before computing `CN`.]

```{r}
# TODO: reproduce the calculation exactly as the original work did it.
# Illustrative example: classical least-squares VIF (car::vif), which
# ignores the GLM's IRLS weights and is therefore not the correct
# transformation for a non-Gaussian GLM.
if (requireNamespace("car", quietly = TRUE)) {
  car::vif(mod)
} else {
  message("The 'car' package is not installed; skipping this reference VIF.")
}
```

[Interpret the result exactly as the original work interpreted it: did
they conclude there was, or was not, problematic collinearity?]

## Diagnostic with multiCollglm (correct transformation)

```{r}
library(multiCollglm)

condition_number(mod)
bkw_diagnostics(mod)
rvif_diagnostics(mod)
```

## Comparison

[Table or text comparing both diagnostics: does the conclusion about the
existence/severity of collinearity change? Which variables are implicated
according to each method?]

| | Original diagnostic | multiCollglm |
|---|---|---|
| Transformation applied | [e.g. none / centered] | Weighted by IRLS, scaled to unit length, without centering |
| Conclusion about collinearity | [...] | [...] |
| Variables implicated | [...] | [...] |

## Conclusion

[2-4 sentences: why the transformation matters in this specific case, and
what would have changed in the original work's conclusions had it been
applied correctly.]

## References

[Cite the original work here with its @key, e.g. `[-@author2024]`, and
add its BibTeX entry to `vignettes/references.bib`.] The two references
below are cited throughout this package's diagnostics [@belsley1980;
@salmeron2025], and will render automatically here once cited above.
