---
title: "Reproduction: plastic plywood manufacturing data (Plastic)"
description: >
  Reproduces the collinearity diagnostic of the plastic plywood
  manufacturing dataset, used by six works fitting different count-data
  GLMs, and documents a citation error (CN = 8634.73) that propagates
  unverified across at least three publications.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

The `Plastic` dataset comes from a statistical-process-control case study
at a medium-sized timber industry that manufactures laminated plastic
plywood (Marcondes Filho and Sant'Anna, 2016). It has 100 reference
observations; the response `Y` is the number of defects per unit area
(count data), and the four regressors are volumetric shrinkage (`X1`, %),
assembly time (`X2`, minutes), wood density (`X3`, g/cm3) and drying
temperature (`X4`, deg C). It is a widely cited dataset in the
collinearity literature because `X1` and `X2` are strongly correlated
with each other, as are `X3` and `X4`.

It has been reused by Sami, Amin and Butt (2022) [`review8`], Tanis and
Asar (2026) [`review19`], Mammadova and Ozkale (2021) [`review27`],
Dawoud (2025) [`review36`], Tanis and Asar (2024) [`review74`] and Ashraf
et al. (2025) [`review81`], each with a different count-data model (Bell,
Conway-Maxwell-Poisson, Poisson, negative binomial).

The original data from Marcondes Filho and Sant'Anna (2016) were verified,
in earlier work in this same project, to be byte-identical (aside from
formatting differences) to the spreadsheet distributed with this package.

## Data

```{r}
library(multiCollglm)
data(Plastic)
str(Plastic)

mod <- glm(Y ~ X1 + X2 + X3 + X4, family = poisson(link = "log"),
           data = Plastic)
summary(mod)
```

## Original diagnostic (a tour of the literature, and a citation error)

Tanis and Asar (2026) and Tanis and Asar (2024) fit a Bell regression
without an intercept, with the data centered and standardized so that
$X'X$ is in correlation form (Tanis and Asar 2024 state this explicitly:
*"The design matrix is centred and standardized so that $X'X$ is in
correlation form"*), obtaining $CN = 74.5281$.

Sami, Amin and Butt (2022) fit a Conway-Maxwell-Poisson model and compute
the condition number as the square root of the eigenvalue ratio of
$X'\hat{W}X$ **excluding the intercept**, with no additional
transformation reported, obtaining $CN = 8634.73$. Ashraf et al. (2025)
report **exactly the same value**, $8634.73$, despite fitting three
different distributions (Conway-Maxwell-Poisson, Poisson and negative
binomial) -- that three different models, necessarily with different IRLS
weight matrices, produce the same condition number to four decimal places
is itself a red flag.

Dawoud (2025) uses only three of the four regressors (assembly time,
density, temperature) with Conway-Maxwell-Poisson and a log link,
obtaining $CN = 13120.14$, with no further detail on the transformation.
Mammadova and Ozkale (2021) report $CN = 2.9558\times10^{5}$ as the ratio
between the largest and smallest eigenvalue of $X'\hat{W}X$, without
specifying a distribution, link, or transformation.

## Diagnostic with multiCollglm

```{r}
methods <- c("RAW", "MP", "MS", "WS", "OZ")

calc_table <- function(mod) {
  do.call(rbind, lapply(methods, function(m) {
    res <- tryCatch(condition_number(mod, method = m),
                     error = function(e) NULL)
    if (is.null(res)) {
      data.frame(method = m, nc_label = NA, condition_number = NA,
                 condition_index = NA)
    } else {
      data.frame(method = m, nc_label = res$nc_label,
                 condition_number = res$condition_number,
                 condition_index = res$condition_index)
    }
  }))
}

cat("All four variables, with intercept:\n")
knitr::kable(calc_table(mod), digits = 4)

mod0 <- glm(Y ~ X1 + X2 + X3 + X4 - 1, family = poisson(link = "log"),
            data = Plastic)
cat("All four variables, without intercept:\n")
knitr::kable(calc_table(mod0), digits = 4)

mod36 <- glm(Y ~ X2 + X3 + X4 - 1, family = poisson(link = "log"),
             data = Plastic)
cat("Dawoud's (2025) subset (X2, X3, X4), without intercept:\n")
knitr::kable(calc_table(mod36), digits = 4)
```

## Comparison

No combination of family/link/transformation available in
`condition_number()` reproduces the three most-cited values in this
literature ($8634.73$; $13120.14$; $2.9558\times10^{5}$), whether with the
four variables or Dawoud's (2025) subset of three, and with or without an
intercept. The values that are obtained span a wide range (from a few
tens to billions depending on the method), but none match the published
figures.

That, together with the fact that Ashraf et al. (2025) report the same
$8634.73$ for three different distributions -- when the IRLS weight
matrix, and hence the condition number, should differ between them --
suggests that $8634.73$ has been **cited from one work to the next
without being independently recalculated**, rather than being a verified
computation in each case. We were unable to establish exactly which
transformation first produced that figure.

Tanis and Asar's value (74.5281, Bell without an intercept, data centered
and standardized into correlation form) could not be independently
verified because Bell regression is not part of R base's `glm()`
families; it is left as reported by the original source.

| | Literature | multiCollglm |
|---|---|---|
| Transformation | Unspecified / "correlation form" without intercept | RAW/MP/MS/WS/OZ, with and without intercept (see tables) |
| Most-cited value | $8634.73$ (identically repeated in 2 works with different distributions) | Not reproducible with any of the 5 definitions |
| Conclusion about collinearity | Severe in all cases | Severe in all cases, though the exact magnitude depends heavily on the chosen method |

## Conclusion

The plastic plywood data show real, well-documented collinearity
(strong correlations between shrinkage and assembly time, and between
density and drying temperature), and that holds under any of the
package's five methods. But the specific figure of $8634.73$ that
dominates this branch of the literature does not hold up: it is not
reproducible from the verified data, and its identical repetition across
three different distributions in Ashraf et al. (2025) is more consistent
with a citation chain than with an independent calculation in each work.
This is the same pattern already seen in other datasets covered by this
package (Nitrogen, Mine): "reference" values that get passed from one
article to the next without ever being recalculated.

## References

- Marcondes Filho, D. and Sant'Anna, A.M.P. (2016). Statistical process
  control applied to a plastic plywood manufacturing process.
- Sami, F., Amin, M. and Butt, M.M. (2022). On the ridge estimation of the
  Conway-Maxwell Poisson regression model with multicollinearity: Methods
  and applications. *Concurrency and Computation: Practice and
  Experience*, 34(1), e6477. <https://doi.org/10.1002/cpe.6477>
- Tanis, C. and Asar, Y. (2026). Almost unbiased Liu estimator in Bell
  regression model. *Communications in Statistics - Simulation and
  Computation*. <https://doi.org/10.1080/03610918.2026.2648869>
- Tanis, C. and Asar, Y. (2024). Almost unbiased ridge estimator in Bell
  regression model: theory and application to plastic polywood data.
  *Statistics*, 58(5), 1031-1045.
  <https://doi.org/10.1080/02331888.2024.2389416>
- Mammadova, U. and Ozkale, M.R. (2021). Profile monitoring for count data
  using Poisson and Conway-Maxwell-Poisson regression-based control
  charts under multicollinearity problem. *Journal of Computational and
  Applied Mathematics*, 388, 113275.
  <https://doi.org/10.1016/j.cam.2020.113275>
- Dawoud, I. (2025). New biased estimators for the Conway-Maxwell-Poisson
  model. *Journal of Statistical Computation and Simulation*, 95(1),
  117-136. <https://doi.org/10.1080/00949655.2024.2421317>
- Ashraf, B., Amin, M., Emam, W., Tashkandy, Y. and Faisal, M. (2025).
  Negative Binomial Regression Model Estimation Using Stein Approach:
  Methods, Simulation, and Applications. *Journal of Mathematics*,
  2025(1), 9134821. <https://doi.org/10.1155/jom/9134821>
