---
title: "Reproduction: Chatterjee and Hadi (1988) -- Nitrogen data"
description: >
  Reproduces the collinearity diagnostic of the Nitrogen dataset (Gamma,
  inverse link), contrasts it against nine published
  reproductions/variants, and shows how each of multiCollglm's five
  condition-number methods (RAW, MP, MS, WS, OZ) responds to this model's
  structural singularity.
bibliography: ../references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

The `Nitrogen` dataset records weather conditions and nitrogen dioxide
concentration (`y`, p.p.h.m.) measured in September 1984 at a monitoring
station in the San Francisco Bay Area, with four regressors: wind speed
(`x1`), maximum temperature (`x2`), insolation (`x3`) and a stability
factor (`x4`). The dataset was introduced by Chatterjee and Hadi [-@chatterjee1988],
and has been reused by at least nine later works to illustrate
collinearity diagnostics in a Gamma GLM with the inverse link.

Chatterjee and Hadi describe the dataset as having 26 observations, but
the original observation numbering jumps from observation 8 to
observation 10 (there is no observation 9): there are in fact only 25
observations, which is what `data(Nitrogen)` includes in this package. As
far as we know, this discrepancy had not previously been flagged in the
literature that reuses this dataset.

## Data

```{r}
library(multiCollglm)
data(Nitrogen)
str(Nitrogen)

mod <- glm(y ~ x1 + x2 + x3 + x4, family = Gamma(link = "inverse"),
           data = Nitrogen)
```

## Diagnostic with multiCollglm and comparison with literature

The first work (within our selection) to use this dataset is Ozkale and Arican [-@ozkalearican2016],
which computes $CN$ on $X'\hat{W}X$ as the
ratio between the largest and smallest eigenvalue (without a square
root), with a reciprocal link and an intercept, **standardizing the four
regressors before adding the column of ones**. They obtain $CN =
213.8097$, which they interpret as evidence of severe collinearity. This
value is then repeated without being recalculated in Amin et al.
[-@amin2020], Lukman et al. [-@lukman2022], Akram et al.
[-@akram2022] and Shewa and Ugwuowo [-@shewa2023]. In the last two
cases the fitted model actually uses a logarithmic link, not the
reciprocal link that originally produced the value, so the $CN$ they cite almost certainly does not correspond to their own weight
matrix.

Kurtoglu [-@kurtoglu2021] performs an analogous calculation but with
the IRLS weights of their own log-gamma model, obtaining a much larger
value: $49,649,929$. Bulut [-@bulut2023] applies the same dataset
under an Inverse Gaussian distribution (not Gamma) with a log link, and
obtains $10,263.65$.

Only two works document a full transformation of the data: Ozkale and
Abbasi [-@ozkaleabbasi2022] weight by $\hat{W}^{1/2}$, center, and rescale
each column to unit length, obtaining $CN = 2.49\times10^{8}$ (and
computing the VIF via a Moore-Penrose pseudoinverse because the resulting
matrix is nearly singular, which they themselves note is why their VIF
values come out artificially low, 0.19-5.49, despite the astronomically
large $CN$). Ozkale [-@ozkale2021], with the same
weighting/centering/unit-length-scaling procedure, reports both an
untransformed scenario ($CN = 1.4820\times10^{4}$) and a transformed one
($CN = 1.5508\times10^{16}$, driven by a numerically zero eigenvalue).

We were unable to reproduce the value from Ozkale and Arican [-@ozkalearican2016]
($\hat\varphi^2 = 0.07572852$, $CN = 213.8097$): refitting the model under
several standardizations consistently yields $\hat\varphi^2 \approx
0.1126$ (a quantity invariant to any rescaling of the design matrix),
which suggests some unreported detail in their estimation procedure
rather than a data or scaling issue.


```{r echo=FALSE}
methods <- c("RAW", "MP", "MS", "WS", "OZ")

tbl <- do.call(rbind, lapply(methods, function(m) {
  res <- tryCatch(condition_number(mod, method = m),
                   error = function(e) NULL)
  if (is.null(res)) {
    data.frame(method = m, nc_label = NA, CN = NA,
               `sqrt(CN)` = NA, check.names = FALSE)
  } else {
    data.frame(method = m, nc_label = res$nc_label,
               CN = res$condition_number,
               `sqrt(CN)` = res$condition_index, check.names = FALSE)
  }
}))
```
```{r}
knitr::kable(tbl, digits = 4)
```


This dataset also illustrates the structural reason already documented in
`?condition_number`: for the Gamma family with its canonical (inverse)
link, `sqrt(w) * eta` equals exactly 1 for every observation, so
centering the IRLS-weighted matrix (as `"OZ"` and `"MS"` do) can leave it
exactly rank-deficient -- with or without an intercept. That is exactly
what explains the near-zero eigenvalues in Ozkale [-@ozkale2021] and
Ozkale and Abbasi [-@ozkaleabbasi2022].


## References
