---
title: "Reproduction: Bodyfat data (TH.data)"
description: >
  Reproduces the collinearity diagnostic of the bodyfat dataset (Gamma,
  log link) used by five works, and documents a structural fact about the
  log link for the Gamma family that explains why two of the package's
  methods coincide exactly in this case.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

The `bodyfat` dataset from the `TH.data` package records body-composition
and anthropometric measurements from 71 healthy German women. The
response, `DEXfat`, is the percentage body fat measured by dual-energy
X-ray absorptiometry (DXA); the regressors include age, waist and hip
circumference, elbow and knee breadth, and several derived anthropometric
variables (`anthro3a`, `anthro3b`, `anthro3c`, `anthro4`). The original
goal is to predict body fat from simpler anthropometric measurements.

Five of the reviewed works use this dataset under a Gamma distribution:
Almuqrin and AbaOud (2025) [`review23`], Al-Ghamdi et al. (2025)
[`review44`], Asar and Korkmaz (2022) [`review51`], Seifollahi et al.
(2024) [`review76`] and El-Masry et al. (2025) [`review127`]. All of them
state that they compute the condition number as the square root of the
ratio between the largest and smallest eigenvalue of $X'\hat{W}X$.

## Data

```{r}
library(multiCollglm)
if (!requireNamespace("TH.data", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
  message("The 'TH.data' package is not installed; skipping this article.")
}
data("bodyfat", package = "TH.data")
str(bodyfat)

mod <- glm(DEXfat ~ age + waistcirc + hipcirc + elbowbreadth + kneebreadth +
             anthro3a + anthro3b + anthro3c + anthro4,
           family = Gamma(link = "log"), data = bodyfat)
summary(mod)
```

## Original diagnostic (a tour of the literature)

Almuqrin and AbaOud (2025), Al-Ghamdi et al. (2025) and El-Masry et al.
(2025) report an identical value, $CN = 3394.30$; none of them explicitly
states which link function was used, so we could not reproduce that
specific value. Asar and Korkmaz (2022) do specify a log link and
untransformed data, and obtain $CN = 3410.63$. Seifollahi et al. (2024)
also claims to use a log link but reports a different value, $CN =
4026.23$ -- an inconsistency within the literature itself, since two
works that claim to follow exactly the same procedure (Gamma, log,
untransformed) do not agree with each other.

## Diagnostic with multiCollglm

```{r}
methods <- c("RAW", "MP", "MS", "WS", "OZ")

tbl <- do.call(rbind, lapply(methods, function(m) {
  res <- tryCatch(condition_number(mod, method = m),
                   error = function(e) NULL)
  if (is.null(res)) {
    data.frame(method = m, nc_label = NA, condition_number = NA,
               condition_index = NA)
  } else {
    data.frame(method = m, nc_label = res$nc_label,
               condition_number = res$condition_number,
               condition_index = res$condition_index)
  }
}))
knitr::kable(tbl, digits = 4)
```

A notable feature of this table: `"MP"` and `"WS"` agree **exactly** (not
just approximately). The reason is structural, specific to this
family/link combination: for a Gamma GLM with a **log** link, the IRLS
weight of every observation at convergence is

$$w_i = \frac{1}{V(\mu_i)}\left(\frac{d\mu_i}{d\eta_i}\right)^2 = \frac{1}{\mu_i^2}\cdot\mu_i^2 = 1,$$

i.e. **constant and equal to 1 for every observation**, regardless of the
fitted values. This can be checked directly:

```{r}
range(mod$weights)
```

With IRLS weights identically equal to 1, weighting by $\hat{W}^{1/2}$
does nothing (it's the same as the unweighted matrix), so rescaling to
unit length *before* fitting (`"MP"`) or *after* fitting (`"WS"`)
produces exactly the same matrix, and hence the same condition number.
This is the log-link analogue, for Gamma, of the identity already
documented for Gamma with the canonical inverse link (`sqrt(w) * eta ≡
1`, see `?condition_number`): every family/link combination has its own
structural identity in the IRLS weights, and it is worth keeping in mind
when interpreting any weighted collinearity diagnostic.

## Comparison

| | Literature | multiCollglm |
|---|---|---|
| Untransformed (`"RAW"`) | $CN = 3410.63$ (Asar and Korkmaz 2022, square root) | `r round(condition_number(mod, method = "RAW")$condition_index, 4)` |
| Rescaled to unit length, no centering | $CN \approx 352.04$ | `r round(condition_number(mod, method = "WS")$condition_index, 4)` |
| Centered + scaled (`"OZ"`) | -- | `r round(condition_number(mod, method = "OZ")$condition_index, 4)` |

Asar and Korkmaz's (2022) value reproduces with `method = "RAW"` to four
decimal places: **3410.6317** versus their **3410.63**. The "consensus"
value from the other three works (3394.30) comes very close but not
exactly, which suggests they used a slightly different link (or some
minor sampling variant) that they did not document; and Seifollahi et
al.'s (2024) value, despite claiming the same procedure as Asar and
Korkmaz, is close to neither.

## Conclusion

Once the already IRLS-weighted matrix is rescaled to unit length without
centering (`"WS"`), the condition number drops from the thousands to a
few hundred: the conclusion of severe collinearity holds, but is no
longer dominated by the disparity in scale between variables as different
as age (in years) and elbow breadth (in centimeters). The more
interesting finding for this dataset, however, is the structural identity
of the IRLS weights under Gamma with a log link: because they are always
equal to 1, two of the package's five methods coincide exactly, which
does not happen in general for other family/link combinations (see the
Nitrogen article, where none of the five methods agree with each other).

## References

- Penrose, K.W., Nelson, A.G. and Fisher, A.G. (1985). Generalized body
  composition prediction equation for men using simple measurement
  techniques. *Medicine & Science in Sports & Exercise*, 17(2), 189.
- Almuqrin, M.A. and AbaOud, M. (2025). Developing a new modified
  two-parameter Liu estimator for the gamma regression model: Method,
  simulation and application to health data. *Alexandria Engineering
  Journal*, 129, 1212-1222. <https://doi.org/10.1016/j.aej.2025.08.033>
- Al-Ghamdi, M.N., Abonazel, M.R., Dawoud, I., Algamal, Z.Y. and Azazy, A.R.
  (2025). A new estimator of the gamma regression model: theory,
  simulation, and application to body fat data. *Communications in
  Mathematical Biology and Neuroscience*, 2025, Article 53.
  <https://doi.org/10.28919/cmbn/9149>
- Asar, Y. and Korkmaz, M. (2022). Almost unbiased Liu-type estimators in
  gamma regression model. *Journal of Computational and Applied
  Mathematics*, 403, 113819. <https://doi.org/10.1016/j.cam.2021.113819>
- Seifollahi, S., Bevrani, H. and Kamary, K. (2024). Inequality restricted
  estimator for gamma regression: Bayesian approach as a solution to the
  multicollinearity. *Communications in Statistics - Theory and Methods*,
  53(23), 8297-8311. <https://doi.org/10.1080/03610926.2023.2281267>
- El-Masry, A.M., Abu-Hamed, A.A., Abonazel, M.R., Omara, T.M. and Khattab,
  I.G. (2025). A new gamma regression estimate of female body fat
  percentage: advanced statistical modeling. *Communications in
  Mathematical Biology and Neuroscience*, 2025, Article 133.
  <https://doi.org/10.28919/cmbn/9493>
