---
title: "Reproduction: Bodyfat data (TH.data)"
description: >
  Reproduces the collinearity diagnostic of the bodyfat dataset (Gamma,
  log link) used by five works, and documents a structural fact about the
  log link for the Gamma family that explains why two of the package's
  methods coincide exactly in this case.
bibliography: ../references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

The `bodyfat` dataset from the `TH.data` package records body-composition
and anthropometric measurements from 71 healthy German women. The
response, `DEXfat`, is the percentage body fat measured by dual-energy
X-ray absorptiometry (DXA); the regressors include age, waist and hip
circumference, elbow and knee breadth, and several derived anthropometric
variables (`anthro3a`, `anthro3b`, `anthro3c`, `anthro4`). The original
goal, following the body-composition prediction approach of Penrose,
Nelson and Fisher [-@penrose1985], is to predict body fat from simpler
anthropometric measurements.

Five of the reviewed works use this dataset under a Gamma distribution:
Almuqrin and AbaOud [-@almuqrin2025], Al-Ghamdi et al.
[-@alghamdi2025], Asar and Korkmaz [-@asar2022],
Seifollahi et al. [-@seifollahi2024] and
El-Masry et al. [-@elmasry2025]. All of them
state that they compute their $CN$ as the square root of the
ratio between the largest and smallest eigenvalue of $X'\hat{W}X$.

## Data

```{r}
library(multiCollglm)
if (!requireNamespace("TH.data", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
  message("The 'TH.data' package is not installed; skipping this article.")
}
data("bodyfat", package = "TH.data")
str(bodyfat)

mod <- glm(DEXfat ~ age + waistcirc + hipcirc + elbowbreadth + kneebreadth +
             anthro3a + anthro3b + anthro3c + anthro4,
           family = Gamma(link = "log"), data = bodyfat)
```

```{r echo=FALSE}

methods <- c("RAW", "MP", "MS", "WS", "OZ")

tbl <- do.call(rbind, lapply(methods, function(m) {
  res <- tryCatch(condition_number(mod, method = m),
                   error = function(e) NULL)
  if (is.null(res)) {
    data.frame(method = m, nc_label = NA, CN = NA,
               `sqrt(CN)` = NA, check.names = FALSE)
  } else {
    data.frame(method = m, nc_label = res$nc_label,
               CN = res$condition_number,
               `sqrt(CN)` = res$condition_index, check.names = FALSE)
  }
}))
```

## Diagnostic with multiCollglm and comparison with literature

Almuqrin and AbaOud [-@almuqrin2025], Al-Ghamdi et al. [-@alghamdi2025]
and El-Masry et al. [-@elmasry2025] report an identical value, $CN =
3394.30$; none of them explicitly
states which link function was used, so we could not reproduce that
specific value. Asar and Korkmaz [-@asar2022] do specify a log link and
untransformed data, and obtain $CN = 3410.63$. Seifollahi et al.
[-@seifollahi2024] also claims to use a log link but reports a different value, $CN =
4026.23$ -- an inconsistency within the literature itself, since two
works that claim to follow exactly the same procedure (Gamma, log,
untransformed) do not agree with each other.

## Diagnostic with multiCollglm

```{r}
knitr::kable(tbl, digits = 4)
```

A notable feature of this table: `"MP"` and `"WS"` agree **exactly** (not
just approximately). The reason is structural, specific to this
family/link combination: for a Gamma GLM with a **log** link, the IRLS
weight of every observation at convergence is

$$w_i = \frac{1}{V(\mu_i)}\left(\frac{d\mu_i}{d\eta_i}\right)^2 = \frac{1}{\mu_i^2}\cdot\mu_i^2 = 1,$$

i.e. **constant and equal to 1 for every observation**, regardless of the
fitted values. This can be checked directly:

```{r}
range(mod$weights)
```

With IRLS weights identically equal to 1, weighting by $\hat{W}^{1/2}$
does nothing (it's the same as the unweighted matrix), so rescaling to
unit length *before* fitting (`"MP"`) or *after* fitting (`"WS"`)
produces exactly the same matrix, and hence the same $CN$.
This is the log-link analogue, for Gamma, of the identity already
documented for Gamma with the canonical inverse link (`sqrt(w) * eta ≡
1`, see `?condition_number`): every family/link combination has its own
structural identity in the IRLS weights, and it is worth keeping in mind
when interpreting any weighted collinearity diagnostic.

Asar and Korkmaz's [-@asar2022] value reproduces with `method = "RAW"` (3410.63). The "consensus"
value from the other three works (3394.30) comes very close but not
exactly, which suggests they used a slightly different link (or some
minor sampling variant) that they did not document; and Seifollahi et
al.'s [-@seifollahi2024] value, despite claiming the same procedure as Asar and
Korkmaz, is close to neither.


## References
