---
title: "Reproduction: plastic plywood manufacturing data (Plastic)"
description: >
  Reproduces the collinearity diagnostic of the plastic plywood
  manufacturing dataset, used by six works fitting different count-data
  GLMs, and documents a citation error (CN = 8634.73) that propagates
  unverified across at least three publications.
bibliography: ../references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

The `Plastic` dataset comes from a statistical-process-control case study
at a medium-sized timber industry that manufactures laminated plastic
plywood [-@marcondes2016]. It has 100 reference
observations; the response `Y` is the number of defects per unit area
(count data), and the four regressors are volumetric shrinkage (`X1`, %),
assembly time (`X2`, minutes), wood density (`X3`, g/cm3) and drying
temperature (`X4`, deg C). It is a widely cited dataset in the
collinearity literature because `X1` and `X2` are strongly correlated
with each other, as are `X3` and `X4`.

It has been reused by Sami, Amin and Butt [-@sami2022], Tanis and
Asar [-@tanis2026], Mammadova and Ozkale [-@mammadova2021],
Dawoud [-@dawoud2025], Tanis and Asar [-@tanis2024] and Ashraf
et al. [-@ashraf2025], each with a different count-data model (Bell,
Conway-Maxwell-Poisson, Poisson, negative binomial).


## Data

```{r}
library(multiCollglm)
data(Plastic)
str(Plastic)

mod <- glm(Y ~ X1 + X2 + X3 + X4, family = poisson(link = "log"),
           data = Plastic)
```

```{r echo=FALSE}
methods <- c("RAW", "MP", "MS", "WS", "OZ")

calc_table <- function(mod) {
  do.call(rbind, lapply(methods, function(m) {
    res <- tryCatch(condition_number(mod, method = m),
                     error = function(e) NULL)
    if (is.null(res)) {
      data.frame(method = m, nc_label = NA, CN = NA,
                 `sqrt(CN)` = NA, check.names = FALSE)
    } else {
      data.frame(method = m, nc_label = res$nc_label,
                 CN = res$condition_number,
                 `sqrt(CN)` = res$condition_index, check.names = FALSE)
    }
  }))
}
```

## Diagnostic with multiCollglm and comparison with literature

Tanis and Asar [-@tanis2026] and Tanis and Asar [-@tanis2024] fit a Bell regression
without an intercept, with the data centered and standardized so that
$X'X$ is in correlation form (Tanis and Asar [-@tanis2024] state this explicitly:
*"The design matrix is centered and standardized so that $X'X$ is in correlation form"*), obtaining $CN = 74.5281$.

Sami, Amin and Butt [-@sami2022] fit a Conway-Maxwell-Poisson model and compute
their $CN$ as the square root of the eigenvalue ratio of
$X'\hat{W}X$ **excluding the intercept**, with no additional
transformation reported, obtaining $CN = 8634.73$. Ashraf et al. [-@ashraf2025]
report **exactly the same value**, $8634.73$, despite fitting three
different distributions (Conway-Maxwell-Poisson, Poisson and negative
binomial) -- that three different models, necessarily with different IRLS
weight matrices, produce the same $CN$ to four decimal places
is itself a red flag.

Dawoud [-@dawoud2025] uses only three of the four regressors (assembly time,
density, temperature) with Conway-Maxwell-Poisson and a log link,
obtaining $CN = 13120.14$, with no further detail on the transformation.
Mammadova and Ozkale [-@mammadova2021] report $CN = 2.9558\times10^{5}$ as the ratio
between the largest and smallest eigenvalue of $X'\hat{W}X$, without
specifying a distribution, link, or transformation.

```{r}
cat("All four variables, with intercept:\n")
knitr::kable(calc_table(mod), digits = 4)

mod1 <- glm(Y ~ X1 + X2 + X3 + X4 - 1, family = poisson(link = "log"),
            data = Plastic)
cat("All four variables, without intercept:\n")
knitr::kable(calc_table(mod1), digits = 4)

mod2 <- glm(Y ~ X2 + X3 + X4 - 1, family = poisson(link = "log"),
             data = Plastic)
cat("Dawoud's (2025) subset (X2, X3, X4), without intercept:\n")
knitr::kable(calc_table(mod2), digits = 4)
```

No combination of family/link/transformation available in
`condition_number()` reproduces the three most-cited values in this
literature ($8634.73$; $13120.14$; $2.9558\times10^{5}$), whether with the
four variables or Dawoud's [-@dawoud2025] subset of three, and with or without an
intercept. The values that are obtained span a wide range (from a few
tens to billions depending on the method), but none match the published
figures.

That, together with the fact that Ashraf et al. [-@ashraf2025] report the same
$8634.73$ for three different distributions (when the IRLS weight
matrix, and hence $CN$, should differ between them)
suggests that $8634.73$ has been **cited from one work to the next
without being independently recalculated**, rather than being a verified
computation in each case. We were unable to establish exactly which
transformation first produced that figure.

Tanis and Asar's [-@tanis2026] value (74.5281, Bell without an intercept, data centered
and standardized into correlation form) could not be independently
verified because Bell regression is not part of R base's `glm()`
families; it is left as reported by the original source.


## References
