---
title: "Reproduction: coal mine roof-fracture data (Mine)"
description: >
  Reproduces the collinearity diagnostic of the coal mine roof-fracture
  dataset, used by five works fitting different count-data GLMs, and
  shows that the "RAW" (untransformed) method exactly reproduces the
  most-cited value in the literature.
bibliography: ../references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Context

The `Mine` dataset records 44 observations of roof fractures in coal
mines in the Appalachian region of West Virginia, with four continuous
predictors: inner burden thickness, percentage extraction of the
previously mined lower seam, lower seam height, and time since the mine
opened. It is the real-data application shared by five reviewed works,
each fitting a different count-data GLM: Kurtoglu and Ozkale
[-@kurtoglu2019a] and Kurtoglu and Ozkale [-@kurtoglu2019b] (same authors,
a restricted ridge and a restricted Liu estimator respectively, both with
Poisson and a log link); Sami, Amin and Butt [-@sami2022]
(Conway-Maxwell-Poisson, a "COMPRR" ridge estimator); Hussein and Algamal
[-@hussein2026] (Bell regression, a Kibria-Lukman-type estimator);
and Sami et al. [-@sami2026] (discrete Weibull, a "DWRE" ridge
estimator).

Sami, Amin and Butt [-@sami2022] state that they analyze a different
distribution but provide no diagnostic measures of their own (which
should differ across distributions), instead citing a previous paper to
claim that the dataset exhibits collinearity. Sami et al. [-@sami2026]
do the same, citing Sami, Amin and Butt [-@sami2022] despite also using a
different distribution; moreover, based on the code in Sami et al.'s
[-@sami2026] appendix, the diagnostic appears to be computed on the
unweighted $X'X$, and although the code includes a standardization step,
it is never actually applied.

## Data

```{r}
library(multiCollglm)
data(Mine)
str(Mine)

mod <- glm(y ~ x1 + x2 + x3 + x4, family = poisson(link = "log"),
           data = Mine)
```
```{r echo=FALSE}
methods <- c("RAW", "MP", "MS", "WS", "OZ")

tbl <- do.call(rbind, lapply(methods, function(m) {
  res <- tryCatch(condition_number(mod, method = m),
                   error = function(e) NULL)
  if (is.null(res)) {
    data.frame(method = m, nc_label = NA, CN = NA,
               `sqrt(CN)` = NA, check.names = FALSE)
  } else {
    data.frame(method = m, nc_label = res$nc_label,
               CN = res$condition_number,
               `sqrt(CN)` = res$condition_index, check.names = FALSE)
  }
}))
```

## Diagnostic with multiCollglm and comparison with literature

`sqrt(CN)` (square-root scale) of `method = "RAW"` -- no
transformation at all, directly on $X'\hat{W}X$ -- is
**`r round(condition_number(mod, method = "RAW")$condition_index, 4)`**,
which matches, up to rounding, the **2093.32** reported by Kurtoglu and
Ozkale [-@kurtoglu2019a; -@kurtoglu2019b].

Hussein and Algamal [-@hussein2026], with a Bell distribution, obtain **2139.89** (same order
of magnitude, consistent with a different but Poisson-like count
distribution for these data).

```{r}
knitr::kable(tbl, digits = 4)
```

This dataset is a good example of how the transformation of data can affect the results in `CN`. Note that the diagnostic's magnitude changes substantially.


## References
